Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-07-20 doc/extrinsic_type_inference_paper/paper.tex
switch to term placeholder

\title{Fully Extrinsic Type Inference (DRAFT 2025-06-17)}
\author{Thomas Logan}

@@

\TODO{lean uses the term "metavariable" for our concept of "plastic variable". Should we conform to Lean's terminology?}

@@

\TODO{We use the term "plastic variable" to avoid the potentially misleading terminology of "meta variable"}

\TODO{Other options are "unknown", "hole", "logic variable", "placeholder", "localetic variable"}

\TODO{replace "plastic" with "placeholder"}

@@

\noindent
For a given context and expression, 
proof typing attempts to infer a type and a subtyping sequence,
such that if the context holds,
then there is a solution to the generated subtyping sequence, under which 
the expression inhabits the inferred type.
The context consists of skolems \ms{\Theta}, subtyping constraints \ms{\Delta}, and typings \ms{\Gamma}. 
The contextual subtyping constraints and typings constrain the search space of valid typings, 
such that the generated subtyping constraints are consistent with them. 
In fact, the generated subtyping constraints contain
all the contextual subtyping constraints, as it keeps the definition a bit less cluttered.
Skolems are variables that further constrain the search space. 
The generated subtyping constraints must be satisfiable under every interpretation of
skolems for which the contextual constraints are satisfiable.  
Proof typing may also generate additional skolems, which can be used to constrain
subsequent sub-problems.
If a variable is not in the set of skolems, then we consider it to be plastic,
in that it can be constrained in any way that is consistent with the other subtyping constraints
and the typing.

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of locales \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each locale into a type,
such that the skolems and plastics are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

@@

\begin{definition} 
  \label{def:proof_subtyping_plastic_elimination}
  \emph{proof subtyping (plastic elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'} 
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \tau' \J{<:} \alpha \in \Delta \implies \tau' \J{<:} \tau \in \Delta^\dagger)
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{+} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    % }
  \end{mathpar}
\end{definition}

\noindent
For plastic elimination \ms{\alpha \subtypes \tau} where \ms{\alpha \notin \Theta}, 
it ensures that the supertype is consistent with all lower bounds of the plastic,
solutions against those lower bounds. Then it learns that the supertype is an upper bound 
of the plastic.

\begin{definition} 
  \label{def:proof_subtyping_plastic_introduction}
  \emph{proof subtyping (plastic introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \alpha \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \alpha \J{<:} \tau' \in \Delta \implies \tau \J{<:} \tau' \in \Delta^\dagger) 
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{-} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails 
    %   \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    % }
  \end{mathpar}
\end{definition}

\noindent
Plastic introduction is the dual of plastic elimination, so its rule
relies on analogous techniques.

@@

\noindent
For skolem introduction \ms{\tau \subtypes \alpha} where \ms{\alpha \in \Theta},
there are two possibilities.
In the first scenario, the skolem has a plastic as a lower bound.
If effect, this enables learning lower bounds on the skolem,
so the solver uses the same techniques as it did for plastic elimination.
I the second scenario, there is a lower bound that's not a plastic,
so the solver ensures that the skolem introduction is weaker
than the constraints by checking that the subtype subtypes the lower bound that
was found. In practice, the solver avoids checking the subtype against the lower bound, 
if the lower bound is a plastic, as that would cause the solver to diverge due to
a mutual dependence between plastic elimination and skolem elimination.

>>>>

\title{Fully Extrinsic Type Inference (DRAFT 2025-06-20)}
\author{Thomas Logan}

@@

\TODO{lean uses the term "metavariable" for our concept of "placeholder variable". Should we conform to Lean's terminology?}

@@

@@

\noindent
For a given context and expression, 
proof typing attempts to infer a type and a subtyping sequence,
such that if the context holds,
then there is a solution to the generated subtyping sequence, under which 
the expression inhabits the inferred type.
The context consists of skolems \ms{\Theta}, subtyping constraints \ms{\Delta}, and typings \ms{\Gamma}. 
The contextual subtyping constraints and typings constrain the search space of valid typings, 
such that the generated subtyping constraints are consistent with them. 
In fact, the generated subtyping constraints contain
all the contextual subtyping constraints, as it keeps the definition a bit less cluttered.
Skolems are variables that further constrain the search space. 
The generated subtyping constraints must be satisfiable under every interpretation of
skolems for which the contextual constraints are satisfiable.  
Proof typing may also generate additional skolems, which can be used to constrain
subsequent sub-problems.
If a variable is not in the set of skolems, then we consider it to be placeholder,
in that it can be constrained in any way that is consistent with the other subtyping constraints
and the typing.

@@

For a function \ms{f},  
it delegates the work to two helpers. 
First, it leverages function typing consolidation 
\ms{\Theta, \Delta, \Gamma \entails f \liftfun \Pi \sim \Xi}
to infer as many types as possible
and consolidates the results into a sequence of locales \ms{\Pi}.
It must consolidate all of the inferred types in order to
ensure that the inferred type of each 
path's pattern does not subtype the type of subsequent paths' patterns. 
The sequence of subtractable types \ms{\Xi} can be ignored, as it is only
exposed for the sake of being used recursively.
Next, it leverage positive packing 
\ms{\textbf{pack}^+(... \entails \Pi) = \tau}
to rewrite each locale into a type,
such that the skolems and placeholders are converted to
existential and universal bindings,
and then combine all the types by compounding intersection.
Packing the path types together ensures that certain
control flows of application can be verified even when  
the best approximation of the type of the argument
is less precise than the pattern of any one path.

@@

\begin{definition} 
  \label{def:proof_subtyping_placeholder_elimination}
  \emph{proof subtyping (placeholder elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'} 
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \tau' \J{<:} \alpha \in \Delta \implies \tau' \J{<:} \tau \in \Delta^\dagger)
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{+} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails \alpha \subtypes \tau \given \Theta', \Delta'\ \alpha\J{<:}\tau
    % }
  \end{mathpar}
\end{definition}

\noindent
For placeholder elimination \ms{\alpha \subtypes \tau} where \ms{\alpha \notin \Theta}, 
it ensures that the supertype is consistent with all lower bounds of the placeholder,
solutions against those lower bounds. Then it learns that the supertype is an upper bound 
of the placeholder.

\begin{definition} 
  \label{def:proof_subtyping_placeholder_introduction}
  \emph{proof subtyping (placeholder introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \alpha \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha \notin \Theta
      \\
      (\forall \tau' \qua \alpha \J{<:} \tau' \in \Delta \implies \tau \J{<:} \tau' \in \Delta^\dagger) 
      \\
      \Theta, \Delta \entails \Delta^\dagger \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    }
    % \inferrule {
    %   \alpha \notin \Theta
    %   \\
    %   \Theta, \Delta \entails \Delta[\alpha\slash\tau]^{-} \backslash \Delta \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails 
    %   \tau \subtypes \alpha \given \Theta', \Delta'\ \tau\J{<:}\alpha
    % }
  \end{mathpar}
\end{definition}

\noindent
Placeholder introduction is the dual of placeholder elimination, so its rule
relies on analogous techniques.

@@

\noindent
For skolem introduction \ms{\tau \subtypes \alpha} where \ms{\alpha \in \Theta},
there are two possibilities.
In the first scenario, the skolem has a placeholder as a lower bound.
If effect, this enables learning lower bounds on the skolem,
so the solver uses the same techniques as it did for placeholder elimination.
I the second scenario, there is a lower bound that's not a placeholder,
so the solver ensures that the skolem introduction is weaker
than the constraints by checking that the subtype subtypes the lower bound that
was found. In practice, the solver avoids checking the subtype against the lower bound, 
if the lower bound is a placeholder, as that would cause the solver to diverge due to
a mutual dependence between placeholder elimination and skolem elimination.
